Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-15 paper.tex
figures updated

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $44$  & $1310$ \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $54$  & $1280$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $83$  & $870$  \\
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $106$ & $730$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement. See Section II A 1.\\
  (2) The boost from substructures have been integrated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (5) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the LP and supersymmetric DM models
used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3
we calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find
promising targets we estimate the fluxes from DM and CRs of all
clusters in the extended HIFLUGCS sample in Section 5. In Section
6 we conclude and discuss our results.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that inverse
  Compton upscattered both CMB photons (green dotted) as well as dust
  and star photons (orange dash-dotted), and CR induced gamma-ray
  emission (red solid). Each panel is associated with an individual DM
  BM model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$. The substructures boost
  the gamma-ray flux from IC upscattered CMB and continuum emission by
  570 while the IC upscattered SD photons is boosted by 20. Upper limits
from Fermi-LAT after 18 months 
 \protect \cite{2010ApJ...717L..71A} are also shown. We find
  bright prospects for detecting the continuum emission, that is
  dominating the total emission in the GeV energy range, in the models
  $\Ip$ and $\Kp$.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.v12.0.1deg.1.6T.SubMass.elmu.SF300.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.BMv12.0.1deg.SubMass.noMW.woGal.eps}
\caption{\it Comparing the flux from different inverse Compton
  upscattered radiation fields. We show the differential inverse
  Compton emission induced by leptophilic DM in the left panel and by
  the $\Kp$ benchmark model in the right panel. The contribution from
  each individual radiation field from top line to bottom line; CMB
  (green dashed-triple-dotted), dust (purple dashed), stars (blue
  dashed-dotted). The sum of the three components are shown with the
  orange solid line.  The red solid line show the CR induced gamma-ray
  flux. The black arrow shows the differential upper limit from Fermi
  \protect \cite{2010ApJ...717L..71A} indicating that the model
  assumptions such as Sommerfeld and/or substructure boost are in
  conflict with the upper limit. We also show projected CTA point
  source sensitivities ($5\sigma$, 50h).  All fluxes are calculated
  for the Fornax cluster within $\rvir$ using a point spread function
  of $0.1\deg$. For this cluster, the boost from Sommerfeld is about
  130 and the enhancement due to substructures from IC upscattered CMB
  and SD photons is 570 and 20, respectively.}
 \label{fig:IR_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/flux.int.v11.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
 \caption{\it Comparing the energy integrated flux from different
   models. We show the emission from; CR induced emission (red
   solid), leptophilic (LP) models that include both final state
   radiation and IC upscattered CMB, dust and starlight (green
   dash-dotted), and benchmark $\Kp$ models (BM) include continuum
   emission, and IC upscattered CMB, dust and starlight (blue
   dashed). The black arrow show the integrated flux upper limit set
   by Fermi-LAT again indicating challenges for the assumptions
   underlying the LP model. The emission is calculated for the Fornax
   cluster using a point spread function of $0.1\deg$. The boost from
   Sommerfeld and substructures is about 130 and 570, respectively. }
 \label{fig:flux_int}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/SB.v12.1GeV.SF300.SubMass.elmu.eps}
 \caption{\it Comparing the intrinsic surface brightness from
   different clusters without taking PSF effects into account. We show
   the CR induced emission (red), leptophilic emission (green), and
   emission from the $\Kp$ benchmark model (blue). The different
   line styles each represent a cluster; Fornax (solid), Coma (dotted),
   Virgo (dashed), and Perseus (dash-dotted). The boost from
   Sommerfeld effect and substructures is about 130 and 570,
   respectively. Notice the very similar shape and normalization of
   the DM models, while the CRs have a much larger scatter in the
   profiles due to the large variation of the gas fraction in these clusters.}
 \label{fig:SB_clu}
\end{figure}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v12.1GeV.SF300.noSuB.vs.SubMass.elmu.eps}
 \caption{\it The impact of substructures on the intrinsic surface
   brightness profile of the Fornax cluster at 1~GeV. The emission
   induced by CRs is denoted by the red solid line, the leptophilic
   models by green lines, and the DM $\Kp$ benchmark model by blue
   lines. The dashed line show the brightness where the boost from
   substructures is included while the substructures are excluded in
   the dotted line. The boost from Sommerfeld and substructures is
   about 130 and 570, respectively. Notice the nearly flat profiles
   when substructures are included, and the relative large boost at
   $0.01\rvir$ that is due to projection of peripheral substructures
   onto the cluster center.}
 \label{fig:SB_sub}
\end{figure}

>>>>

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters within $\rvir$.}
\begin{tabular}{l  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $60$  & $1040$ \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $70$  & $910$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $105$  & $680$  \\
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $130$  & $580$  \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement. See Section II A 1.\\
  (2) The boost from substructures have been integrated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (5) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

This work studies the prospects for detecting extended gamma-ray
emission from galaxy clusters induced by both DM as well as CR
protons. In Section 2 we discuss the theory of DM and gamma-rays. In
particular, we focus on the leptophilic (LP) and supersymmetric DM
models used in this work, and outline the framework for estimating the
gamma-ray emission from various radiative processes. In Section 3 we
calculate the gamma-ray fluxes and spectral distributions from four
clusters and compare it to upper limits set by Fermi-LAT. For the same
clusters we derive the gamma-ray surface brightness profiles in
Section 4. To extend the analysis to more clusters and find promising
targets we estimate the fluxes from DM and CRs of all clusters in the
extended HIFLUGCS sample in Section 5. In Section 6 we conclude and
discuss our results.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v12.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that inverse
  Compton upscattered both CMB photons (green dotted) as well as dust
  and star photons (orange dash-dotted), and CR induced gamma-ray
  emission (red solid). Each panel is associated with an individual DM
  BM model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$. The substructures boost
  the gamma-ray flux from IC upscattered CMB and continuum emission by
  730 while the IC upscattered SD photons is boosted by 20. Upper
  limits from Fermi-LAT after 18 months \protect
  \cite{2010ApJ...717L..71A} are also shown. We find bright prospects
  for detecting the continuum emission, that is dominating the total
  emission in the GeV energy range, in the models $\Ip$ and $\Kp$.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.v12.0.1deg.1.6T.SubMass.elmu.SF300.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.BMv12.0.1deg.SubMass.noMW.woGal.eps}
\caption{\it Comparing the flux from different inverse Compton
  upscattered radiation fields. We show the differential inverse
  Compton emission induced by leptophilic DM in the left panel and by
  the $\Kp$ benchmark model in the right panel. The contribution from
  each individual radiation field from top line to bottom line; CMB
  (green dashed-triple-dotted), dust (purple dashed), stars (blue
  dashed-dotted). The sum of the three components are shown with the
  orange solid line.  The red solid line show the CR induced gamma-ray
  flux. The black arrow shows the differential upper limit from Fermi
  \protect \cite{2010ApJ...717L..71A} indicating that the model
  assumptions such as Sommerfeld and/or substructure boost are in
  conflict with the upper limit. We also show projected CTA point
  source sensitivities ($5\sigma$, 50h).  All fluxes are calculated
  for the Fornax cluster within $\rvir$ using a point spread function
  of $0.1\deg$. For this cluster, the boost from Sommerfeld is about
  130 and the enhancement due to substructures from IC upscattered CMB
  and SD photons is 730 and 20, respectively.}
 \label{fig:IR_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/flux.int.v12.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
 \caption{\it Comparing the energy integrated flux from different
   models. We show the emission from; CR induced emission (red
   solid), leptophilic (LP) models that include both final state
   radiation and IC upscattered CMB, dust and starlight (green
   dash-dotted), and benchmark $\Kp$ models (BM) include continuum
   emission, and IC upscattered CMB, dust and starlight (blue
   dashed). The black arrow show the integrated flux upper limit set
   by Fermi-LAT again indicating challenges for the assumptions
   underlying the LP model. The emission is calculated for the Fornax
   cluster using a point spread function of $0.1\deg$. The boost from
   Sommerfeld and substructures is about 130 and 730, respectively. }
 \label{fig:flux_int}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/SB.v12.1GeV.SF300.SubMass.elmu.eps}
 \caption{\it Comparing the intrinsic surface brightness from
   different clusters without taking PSF effects into account. We show
   the CR induced emission (red), leptophilic emission (green), and
   emission from the $\Kp$ benchmark model (blue). The different line
   styles each represent a cluster; Fornax (solid), Coma (dotted),
   Virgo (dashed), and Perseus (dash-dotted). \C{We include the boost
   from both substructures and Sommerfeld effect.} Notice the very
   similar shape and normalization of the DM models, while the CRs
   have a much larger scatter in the profiles due to the large
   variation of the gas fraction in these clusters.}
 \label{fig:SB_clu}
\end{figure}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v12.1GeV.SF300.noSuB.vs.SubMass.elmu.eps}
 \caption{\it The impact of substructures on the intrinsic surface
   brightness profile of the Fornax cluster at 1~GeV. The emission
   induced by CRs is denoted by the red solid line, the leptophilic
   models by green lines, and the DM $\Kp$ benchmark model by blue
   lines. The dashed line show the brightness where the boost from
   substructures is included while the substructures are excluded in
   the dotted line. The boost from Sommerfeld and substructures is
   about 130 and 730, respectively. Notice the nearly flat profiles
   when substructures are included, and the relative large boost at
   $0.01\rvir$ that is due to projection of peripheral substructures
   onto the cluster center.}
 \label{fig:SB_sub}
\end{figure}
